\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-02-05 analysis/paper/paper.tex
add intro and edit some of the sections

The purpose of the \pkg{casebase} package is to provide practitioners
with an easy-to-use software tool to predict the risk (or cumulative
incidence (CI)) of an event, for a particular patient. The following
points should be noted:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  hazard ratios
\item
  If, however, the absolute risks are of interest, they have to be
  recovered using the semi-parametric Breslow estimator
\item
  Alternative approaches for fitting flexible hazard models for
  estimating absolute risks, not requiring this two-step approach? Yes!
  \citep{hanley2009fitting}
\end{enumerate}

\citep{hanley2009fitting} propose a fully parametric hazard model that
can be fit via logistic regression. From the fitted hazard function,
cumulative incidence and, thus, risk functions of time, treatment and
profile can be easily derived.

@@

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest performing case-base sampling \emph{uniformly}, i.e.~to sample
the base series uniformly from the study base. In terms of Poisson
processes, this sampling strategy corresponds essentially to a
time-homogeneous Poisson process with intensity equal to \(b/B\), where
\(b\) is the number of sampled observations in the base series, and
\(B\) is the total population-time for the study base. More complex
examples are also available; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the probabilities of the sampling
mechanism are proportional to the cardiovascular disease event rate
given by the Framingham score.

The \texttt{casebase} package fits the family of hazard functions of the
form

\[ h(t;X) = \exp[g(t;X)] \] where \(t\) denotes time and \(X\), the
individual's covariate profile. Different functions of \(t\) lead to
different parametric hazard models. The simplest of these models is the
one-parameter exponential distribution which is obtained by taking the
hazard function to be constant over the range of \(t\).

\[ h(t;X) = \exp(\beta_0 + \beta_1 X) \]

The instantaneous failure rate is independent of \(t\), so that the
conditional chance of failure in a time interval of specified length is
the same regardless of how long the individual has been in the study.
This is also known as the \emph{memoryless property} (Kalbfleisch and
Prentice, 2002).

@@

\[ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) \]

Use of \(\log(t)\) yields the Weibull hazard which allows for a power
dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

\[ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) \]

@@

where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
be expected, this functional form is similar to the terms appearing in
the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R+                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]), 
R+      yscale = 100)
R> legend("topleft", 
R+        legend = c("Control group", "Screening group"), 
R+        col = c("red","blue"),
R+        lty = c(1, 1), 
R+        bg = "gray90")
\end{CodeInput}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      yscale = 100)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      yscale = 100)
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R+                                     time = seq(0,15,0.1), 
R+                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      yscale = 100)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

@@

In the next case study, we show how case-base sampling can also be used
in the context of a competing risk analysis. For illustrative purposes,
we will use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is also available as part of the
\pkg{casebase} package.

@@

\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.71 & (0.41, 1.24) & 0.68 & (0.39, 1.19)\\
Disease & 0.55 & (0.31, 1) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 4.30 & (2, 9.27) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.56 & (0.52, 4.7) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

We will compare our estimates to that obtained from a corresponding
Fine-Gray model \citep{fine1999proportional}. The Fine-Gray model is a
semiparametric model for the cause-specific \emph{subdistribution},
i.e.~the function \(f_k(t)\) such that
\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
model allows to directly assess the effect of a covariate on the
subdistribution, as opposed to the hazard. For the computation, we will
use the \pkg{timereg} package \citep{timereg}.

@@

Although the previous case studies provide a broad overview of the
capabilities of \pkg{casebase}, they all have two properties in common:

\begin{itemize}
  \item All covariates are fixed, i.e. they do not change over time;
  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
\end{itemize}

Case-base sampling has been used in the literature to study vaccination
safety, where the exposure period was defined as the week following
vaccination \citep{saarela2015case}. Hence, the main covariate of
interest, i.e.~exposure to the vaccine, was changing over time. In this
context, case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

The purpose of the next case study is to show how \pkg{casebase} can
also be used for survival analysis problems that do not share the
properties above. To this end, we will use the Stanford Heart Transplant
Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
use the version available in the \pkg{surivival} package.

Recall the setting of this study: patients were admitted to the Stanford
program after meeting with their physician and determining that they
were unlikely to respond to other forms of treatment. After enrollment,
the program searches for a suitable donor for the patient, which can
take anywhere between a few days to almost a year. We are interested in
the effect of a heart transplant on survival; therefore, the patient is
considered exposed only after the transplant has occurred.

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    803    760    733    716
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>         
#> 365 0.29
\end{CodeOutput}

@@

\begin{CodeOutput}
#> [1] 0.21
\end{CodeOutput}
\end{CodeChunk}

@@

In this article, we presented the \proglang{R} package \pkg{casebase}
that provides functions to fit smooth parametric hazards and estimate
cumulative incidence functions using case-base sampling. We outlined the
theoretical underpinnings of the approach, we provided details about our
implementation, and we illustrated the merits of the approach and the
package through four case studies.

As a methodological framework, case-base sampling is very flexible. This
flexibility has been explored before in the literature: for example,
Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
to model a time-dependent exposure variable in a vaccine safety study.
As another example, Saarela and Arjas \citeyearpar{saarela2015non}
combined case-base sampling and a Bayesian non-parametric framework to
compute individualized risk assessments for chronic diseases. In the
case studies above, we explored this flexibility along two fronts. On
the one hand, we showed how splines could be used as part of the linear
predictor to model the effect of time on the hazard. This strategy
yielded estimates of the survival function that were qualitatively
similar to semiparametric estimates derived from Cox regression;
however, case-base sampling led to estimates that \emph{varying smoothly
in time}. On the other hand, we also displayed the flexibility of
case-base sampling by showing how it could be combined with penalized
logistic regression to perform variable selection. Similarly, our
package \pkg{casebase} provides the user with the functionality of
Generalized Additive Models. Moreover, case-base sampling can be
combined with quasi-likelihood estimation to fit survival models that
potentially account for the presence of over-dispersion. All of the
examples illustrate how the case-base sampling framework in general, and
the package \pkg{casebase} in particular, allows the user to fit a broad
family of survival functions.

As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
case-base sampling is comprised of three steps: 1) sampling a case
series and a base series from the study; 2) fit the log-hazard as a
linear function of predictors (including time); and 3) use the fitted
hazard to estimate the cumulative incidence curve. Accordingly, our
package provides functions for each step. Moreover, the simple interface
of the \code{fittingSmoothHazard} function resembles the \code{glm}
interface, which should look familiar to new users. Our modular approach
also provides a convenient way to extend our package for new sampling or
fitting strategies.

In the case studies above, we compared the performance of case-base
sampling with that of Cox regression and Fine-Gray models. In terms of
function interface, \pkg{casebase} uses a formula interface that is
closer to that of \code{glm}, in that the event variable is the only
variable appearing on the left-hand side of the formula. By contrast,
both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as
the left-hand side that captures both the event type and time. Both
approaches to modeling yield user-friendly code. In terms of output,
both approaches differ significantly. Case-base sampling produces smooth
hazard and smooth cumulative incidence curves, whereas Cox regression
and Fine-Gray models produce step-wise cumulative incidence functions
and never explicitly model the hazard function. Qualitatively, we showed
that by using splines in the linear predictor, all three models yielded
similar curves. However, the smooth nature of the output of
\pkg{casebase} improves interpretation for consumers of these
predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
comparison between the Cox model and case-base sampling.

@@

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
may want to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools would then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g. \pkg{SmoothHazard},
\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
Again, we hope to include any resulting methodology as part of this
package.

In future versions of the package, we also want to increase the
complement of diagnostic and inferential tools that are currently
available. For example, we would like to include the ability to compute
confidence intervals for the cumulative incidence curve. The delta
method or parametric bootstrap are two different strategies we can use
to construct approximate confidence intervals. Furthermore, we would
like to include functions to compute calibration and discrimination
statistics (e.g.~Brier score, AUC) for our models. Saarela and Arjas
\citeyearpar{saarela2015non} also describe how to obtain a posterior
distribution for the AUC from their model.

@@

This report was generated on 2020-01-28 14:29:58 using the following
computational environment and dependencies:

@@

\begin{CodeInput}
R> # which R packages and versions?
R> devtools::session_info()
\end{CodeInput}

\begin{CodeOutput}
#> - Session info ---------------------------------------------------------------
#>  setting  value                       
#>  version  R version 3.6.2 (2019-12-12)
#>  os       Pop!_OS 19.10               
#>  system   x86_64, linux-gnu           
#>  ui       X11                         
#>  language en_US:en                    
#>  collate  en_US.UTF-8                 
#>  ctype    en_US.UTF-8                 
#>  tz       America/Toronto             
#>  date     2020-01-28                  
#> 
#> - Packages -------------------------------------------------------------------
#>  package     * version    date       lib
#>  assertthat    0.2.1      2019-03-21 [1]
#>  backports     1.1.5      2019-10-02 [1]
#>  broom         0.5.4      2020-01-27 [1]
#>  callr         3.4.1      2020-01-24 [1]
#>  casebase    * 0.2.1.9001 2020-01-27 [1]
#>  cellranger    1.1.0      2016-07-27 [1]
#>  cli           2.0.1      2020-01-08 [1]
#>  codetools     0.2-16     2018-12-24 [4]
#>  colorspace    1.4-1      2019-03-18 [1]
#>  cowplot     * 1.0.0      2019-07-11 [1]
#>  crayon        1.3.4      2017-09-16 [1]
#>  data.table    1.12.8     2019-12-09 [1]
#>  DBI           1.1.0      2019-12-15 [1]
#>  dbplyr        1.4.2      2019-06-17 [1]
#>  desc          1.2.0      2018-05-01 [1]
#>  devtools      2.2.1      2019-09-24 [1]
#>  digest        0.6.23     2019-11-23 [1]
#>  dplyr       * 0.8.3      2019-07-04 [1]
#>  ellipsis      0.3.0      2019-09-20 [1]
#>  evaluate      0.14       2019-05-28 [1]
#>  fansi         0.4.1      2020-01-08 [1]
#>  farver        2.0.3      2020-01-16 [1]
#>  forcats     * 0.4.0      2019-02-17 [1]
#>  foreach       1.4.7      2019-07-27 [1]
#>  fs            1.3.1      2019-05-06 [1]
#>  generics      0.0.2      2018-11-29 [1]
#>  ggplot2     * 3.2.1      2019-08-10 [1]
#>  glmnet      * 3.0-2      2019-12-11 [1]
#>  glue        * 1.3.1      2019-03-12 [1]
#>  gtable        0.3.0      2019-03-25 [1]
#>  haven         2.2.0      2019-11-08 [1]
#>  hms           0.5.3      2020-01-08 [1]
#>  htmltools     0.4.0      2019-10-04 [1]
#>  httr          1.4.1      2019-08-05 [1]
#>  iterators     1.0.12     2019-07-26 [1]
#>  jsonlite      1.6        2018-12-07 [1]
#>  kableExtra  * 1.1.0      2019-03-16 [1]
#>  knitr         1.27       2020-01-16 [1]
#>  labeling      0.3        2014-08-23 [1]

>>>>

Survival analysis has been greatly influenced over the last 50 years by
the partial likelihood approach of the Cox proportional hazard model
\citep{cox1972regression}. This approach provides a flexible way of
assessing the influence of covariates on the hazard function, without
the need to specify a parametric survival model. This flexibility comes
at the cost of decoupling the baseline hazard from the effect of the
covariates. To recover the whole survival curve---or the cumulative
incidence function (CIF)---we then need to separately estimate the
baseline hazard \citep{breslow1972discussion}. This in turn often leads
to stepwise estimates of the survival function.

From the perspective of clinicians and their patients, the most relevant
quantity is often the 5- or 10-year risk of having a certain event given
the patient's particular circumstances, and not the hazard ratio between
a treatment and control group. Therefore, to make sound clinical
decisions, it is important to accurately estimate the \emph{full} hazard
function, which can then be used to estimate the cumulative incidence
function (CIF). From this perspective, the CIF and the survival function
estimates are now smooth functions of time.

With the goal of fitting smooth-in-time hazard functions, Hanley \&
Miettinen \citeyearpar{hanley2009fitting} proposed a general framework
for estimating fully parametric hazard models via logistic regression.
Their approach provides users that are familiar with generalized linear
models with a natural way of fitting parametric survival models.
Moreover, their framework is very flexible: general functions of time
can be estimated (e.g.~using splines or general additive models), and
hence these models retain some of the flexibility of partial likelihood
approaches.

In this article, we present an \proglang{R} package that combines the
ideas of Hanley \& Miettinen into a simple interface. The purpose of the
\pkg{casebase} package is to provide practitioners with an easy-to-use
software tool to predict the risk (or cumulative incidence) of an event,
conditional on a particular patient's covariate profile. Our package
retains the flexibility of case-base sampling and the familiar interface
of the \code{glm} function.

In what follows, we first recall some theoretical details on case-base
sampling and its use for estimating parameteric hazard functions. We
then give a short review of existing \proglang{R} packages that
implement comparable features as \pkg{casebase}. Next, we provide some
details about the implementation of case-base sampling in our package,
and we give a brief survey of its main functions. This is followed by
four case studies that illustrate the flexibility and capabilities of
\pkg{casebase}. We show how the same framework can be used for competing
risk analyses, penalized estimation, and for studies with time-dependent
exposures. Finally, we end the article with a discussion of the results
and of future directions.

@@

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest sampling the base series \emph{uniformly} from the study base.
In terms of Poisson processes, this sampling strategy corresponds
essentially to a time-homogeneous Poisson process with intensity equal
to \(b/B\), where \(b\) is the number of sampled observations in the
base series, and \(B\) is the total population-time for the study base.
More complex examples are also available; see for example Saarela \&
Arjas \citeyearpar{saarela2015non}, where the probabilities of the
sampling mechanism are proportional to the cardiovascular disease event
rate given by the Framingham score.

Let \(g(t; X)\) be the linear predictor such that
\(\log(h(t;X)) = g(t; X)\). Different functions of \(t\) lead to
different parametric hazard models. The simplest of these models is the
one-parameter exponential distribution which is obtained by taking the
hazard function to be constant over the range of \(t\):

\[ \log(h(t; X)) = \beta_0 + \beta_1 X.\]

The instantaneous failure rate is independent of
\(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

\[ \log(h(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.\]

Use of \(\log(t)\) yields the Weibull hazard which allows for a power
dependence of the hazard on time \citep{kalbfleisch2011statistical}:

\[ \log(h(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \]

@@

where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
be expected, this functional form is similar to the terms appearing in
the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R+                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> legend("topleft", 
R+        legend = c("Control group", "Screening group"), 
R+        col = c("red","blue"),
R+        lty = c(1, 1), 
R+        bg = "gray90")
\end{CodeInput}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R+                                     time = seq(0,15,0.1), 
R+                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

@@

In the next case study, we show how case-base sampling can be used in
the context of a competing risk analysis. For illustrative purposes, we
will use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is also available as part of the
\pkg{casebase} package.

@@

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.71 & (0.41, 1.24) & 0.68 & (0.39, 1.19)\\
Disease & 0.55 & (0.31, 1) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 4.30 & (2, 9.27) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.56 & (0.52, 4.7) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

We will compare our estimates to that obtained from a corresponding
Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
is a semiparametric model for the cause-specific \emph{subdistribution},
i.e.~the function \(f_k(t)\) such that
\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
model allows to directly assess the effect of a covariate on the
subdistribution, as opposed to the hazard. For the computation, we will
use the \pkg{timereg} package \citep{timereg}:

@@

In the previous case studies, we only considered covariates that were
fixed at baseline. In this next case study, we will use the Stanford
Heart Transplant data
\citep[\citet{crowley1977covariance}]{clark1971cardiac} to show how
case-base sampling can also be used in the context of time-varying
covariates. As an example that already appeared in the literature,
case-base sampling was to study vaccination safety, where the exposure
period was defined as the week following vaccination
\citep{saarela2015case}. Hence, the main covariate of interest,
i.e.~exposure to the vaccine, was changing over time. In this context,
case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

Recall the setting of Stanford Heart Transplant study: patients were
admitted to the Stanford program after meeting with their physician and
determining that they were unlikely to respond to other forms of
treatment. After enrollment, the program searched for a suitable donor
for the patient, which could take anywhere between a few days to almost
a year. We are interested in the effect of a heart transplant on
survival; therefore, the patient is considered exposed only after the
transplant has occurred.

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    795    760    729    724
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>         
#> 365 0.27
\end{CodeOutput}

@@

\begin{CodeOutput}
#> [1] 0.19
\end{CodeOutput}
\end{CodeChunk}

@@

In this article, we presented the \proglang{R} package \pkg{casebase},
which provides functions for fitting smooth parametric hazards and
estimating cumulative incidence functions using case-base sampling. We
outlined the theoretical underpinnings of the approach, we provided
details about our implementation, and we illustrated the merits of the
approach and the package through four case studies.

As a methodological framework, case-base sampling is very flexible. This
flexibility has been explored before in the literature: for example,
Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
to model a time-dependent exposure variable in a vaccine safety study.
As another example, Saarela and Arjas \citeyearpar{saarela2015non}
combined case-base sampling and a Bayesian non-parametric framework to
compute individualized risk assessments for chronic diseases. In the
case studies above, we explored this flexibility along two fronts. On
the one hand, we showed how splines could be used as part of the linear
predictor to model the effect of time on the hazard. This strategy
yielded estimates of the survival function that were qualitatively
similar to semiparametric estimates derived from Cox regression;
however, case-base sampling led to estimates of the survival function
that \emph{vary smoothly in time}. On the other hand, we also displayed
the flexibility of case-base sampling by showing how it could be
combined with penalized logistic regression to perform variable
selection. Even though we did not illustrate it in this article,
case-base sampling can also be combined with the framework of
\emph{generalized additive models}. This functionality has been
implemented in our package \pkg{casebase}. Similarly, case-base sampling
can also be combined with quasi-likelihood estimation to fit survival
models that can account for the presence of over-dispersion. All of
these examples illustrate how the case-base sampling framework in
general, and the package \pkg{casebase} in particular, allows the user
to fit a broad and flexible family of survival functions.

As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
case-base sampling is comprised of three steps: 1) sampling a case
series and a base series from the study; 2) fit the log-hazard as a
linear function of predictors (including time); and 3) use the fitted
hazard to estimate the cumulative incidence curve. Accordingly, our
package provides functions for each step. Moreover, the simple interface
of the \code{fittingSmoothHazard} function resembles the \code{glm}
interface. This interface should look familiar to new users. Our modular
approach also provides a convenient way to extend our package for new
sampling or fitting strategies.

In the case studies above, we compared the performance of case-base
sampling with that of Cox regression and Fine-Gray models. In terms of
function interface, \pkg{casebase} uses a formula interface that is
closer to that of \code{glm}, in that the event variable is the only
variable appearing on the left-hand side of the formula. By contrast,
both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
that captures both the event type and time. Both approaches to modeling
yield user-friendly code. However, in terms of output, both approaches
differ significantly. Case-base sampling produces smooth hazards and
smooth cumulative incidence curves, whereas Cox regression and Fine-Gray
models produce step-wise cumulative incidence functions and never
explicitly model the hazard function. Qualitatively, we showed that by
using splines in the linear predictor, all three models yielded similar
curves. However, the smooth nature of the output of \pkg{casebase}
provides a more intuitive interpretation for consumers of these
predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
comparison between the Cox model and case-base sampling.

@@

Future work will look at some of the methodological extensions of
case-base sampling. First, to assess the quality of the model fit, we
want to study the properties of the residuals (e.g.~Cox-Snell,
martingale). More work needs to be done to understand these residuals in
the context of the partial likelihood underlying case-base sampling. The
resulting diagnostic tools would then be integrated in this package.
Also, we are interested in extending case-base sampling to account for
interval censoring. This type of censoring is very common in
longitudinal studies, and many packages (e.g. \pkg{SmoothHazard},
\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
Again, we hope to include any resulting methodology as part of this
package.

In future versions of the package, we also want to increase the
complement of diagnostic and inferential tools that are currently
available. For example, we would like to include the ability to compute
confidence intervals for the cumulative incidence curve. The delta
method or parametric bootstrap are two different strategies we can use
to construct approximate confidence intervals. Furthermore, we would
like to include functions to compute calibration and discrimination
statistics (e.g.~Brier score, AUC) for our models. Saarela and Arjas
\citeyearpar{saarela2015non} also describe how to obtain a posterior
distribution for the AUC from their model. Their approach could also be
included in \pkg{casebase}.

@@

This report was generated on 2020-02-05 17:43:41 using the following
computational environment and dependencies:

@@

\begin{CodeInput}
R> # which R packages and versions?
R> # devtools::session_info()
R> sessionInfo()
\end{CodeInput}
